---
title: "Evictions v. Homelessness Analysis"
subtitle: "Sarah Jiang"
output: html_notebook
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, error = FALSE, message = FALSE, warning = FALSE, collapse=TRUE)
```

------------------------------------------------------------------------

## Panel Dataset

| variable     | label                                                                        |
|------------------|------------------------------------|
| state_fips   | state FIPS                                                                   |
| state_name   | state name                                                                   |
| state_abbrev | state abbreviation                                                           |
| year         | year of observed homeless rate                                               |
| fileyear     | = year - 1; year of observed filing rate                                     |
| filerate     | Evictions filing rate per 1,000 renting households.                          |
| homelesscap  | Rate of homelessness per 1,000 residents.                                    |
| chroniccap   | Chronic homelessness rate per 1,000 residents.                               |
| red          | = 1 if proportion of Republican wins \> 60% for Senate elections 1976 - 2020 |
| blue         | = 1 if proportion of Democratic wins \> 60% for Senate elections 1976 - 2020 |
| densityrank  | 2010 Census historical population density ranking                            |
| dense        | = 1 if 2010 Census population density ranking is 26 or lower                 |

```{r load-libs, collapse=TRUE, hide = TRUE}
library(haven)
library(MASS)
library(ggplot2)
data = read_dta("homelesspanel.dta")
```

## Summarize, explore & visualize data

The dataset comprises of 452 observations of eviction filings per 1,000
renting households and number of homeless people per 1,000 residents per
state per year from 2008 to 2019. Not all states and years are
represented. Variables red, blue, densityrank and dense are for the
purpose of profiling states and do not vary within each state.

The variable fileyear = year - 1 with the purpose of reflecting that the effects of evictions may not have an immediately observable effect on the state's population of homeless, therefore we stagger year of filing rate and year of observed rate of homelessness by 1.

```{r}
head(data)
sumstats = summary(data[c('filerate', 'homelesscap', 'chroniccap', 'red', 'blue')])
print(sumstats)
```

Let's take an initial look at the distribution of our main variables of interest, homelesscap and filerate.

```{r}
# create function for consistent aesthetic theme for distribution visualizations.
plot_template = function(data, type = "box", title, label, color = 'seashell') {
  ## create named list of boxplot func & hist func
  plot_types = list(
    box = function(dt) {
      boxplot(dt, breaks = 40, 
              main = title, 
              xlab = label, 
              col = color, 
              cex.main = 0.7, 
              cex.lab = 0.7)},
    hist = function(dt) {
      hist(dt, breaks = 40, 
              main = title, 
              xlab = label, 
              col = color, 
              cex.main = 0.7, 
              cex.lab = 0.7)}
  )
  # choose specified type of plot by indexing plot_types
  plotted = plot_types[[type]]
  plotted(data)
}
```


```{r}
hlabel = 'homeless persons per 1,000 residents'
flabel = 'eviction filings per 1,000 renting households'

# side by side histograms
par(mfrow = c(1, 2))

plot_template(data$homelesscap, type = 'hist',
              title = "Homelessness by state by year (homelesscap)",
              label = hlabel)

plot_template(data$filerate, type = 'hist',
              title = 'Rate of eviction filings by state by year (filerate)',
              label = flabel,
              color = 'lavender')
```
```{r}
# boxplots
par(mfrow = c(1, 2))

plot_template(data$homelesscap,
              title = "Homelessness by state by year (homelesscap)",
              label = hlabel)

plot_template(data$filerate,
              title = 'Rate of eviction filings by state by year (filerate)',
              label = flabel,
              color = 'lavender')
```

The overall distributions of homelesscap and filerate are both very
right-skewed and likely multimodal due to states with unique
circumstances (the high outlier on filing rates is Maryland). 

$~$

## Exploring homelesscap v. filerate

Is there any relationship between eviction filing rates and
homelessness?

```{r}
# create scatterplot of variables of interest with trend lines and CIs
base = ggplot(data = data, mapping = aes(x = filerate, y = homelesscap))

base + geom_point(size = 0.8) + 
  geom_smooth(method = 'lm', color = 'black', linewidth = .7) +
  geom_smooth(color = 'green', linewidth = .7) +
  labs(title = 'Rates of evictions filings & rates of homelessness by state by year') +
  ylab('# homeless persons per 1,000 residents') +
  xlab('# eviction filings per 1,000 renting households')

```

A linear model (illustrated by the straight black line) would clearly not be a good fit for the distribution of our data. It would also violate assumptions regarding heteroskedasticity and the Conditional Mean Assumption, $E(u|x) = 0$. As filerate increases, the variance of the errors decreases. $E(u|x)$ fluctuates as $x$, or filerate, changes, because as filerate increases, we know that $E(u)$ will first become negative around $150 < x < 300$ before becoming positive when $x > 350$. 

$~$

### Initial simple linear regression

$$\hat{homelesscap}_i = \hat\beta_0 + \hat\beta_1{filerate}_i + u_i$$

```{r}
l_model = lm(homelesscap ~ filerate, data = data)
summary(l_model)

# Plot model diagnostics
plot_diagnostics = function(model, x_var = data$filerate, xlabel = 'eviction filings'){
  par(mfrow=c(1, 2))
  plot(x_var, model$residuals, main = 'scatterplot of x_i and u_i', xlab = xlabel, ylab = "residuals")
  plot(model, which = 2)
}

plot_diagnostics(model = l_model)
```

This model doesn't do  great job of explaining homelesscap, having a very low $R^2$ = 0.01456--variation of filing rate only explains around 1.4% of variation in homelesscap. filerate's coefficient of -0.00169 indicates a very slightly negative relationship between filerate and homelesscap that is significant with 95% confidence, with a p-value of 0.0102. The intercept says that on average, a state with no eviction filings will have approximately 1.7885 homeless per 1,000 residents and is highly statistically significant. However, this model violates basically all of our classical linear regression model assumptions:

1. $E(u|x) = 0$: violated, as seen in the scatterplot of $x_i$ and $u_i$, average error is not consistently 0, and the QQ plot shows that they are not Normally distributed
2. No heteroskedasticity: violated; variance of errors is very high at low values of x and decreases when x increases
3. No autocorrelation: violated; errors systematically change as x increases

I will try transforming the data using a log-log model to see if the fit improves.

$~$

### Log-log regression

$$\hat{ln(homelesscap)}_i = \hat\beta_0 + \hat\beta_1{ln(filerate)}_i + \hat\beta_2{red}_i + \hat\beta_3{blue}_i + \hat\beta_4{densityrank}_i + u_i$$

We have a couple of zeros in filerate, so we'll need to drop those corresponding observations first.

```{r collapse = TRUE}
# subset data
data_nozero = subset(data, (filerate>0) & (homelesscap>0))

# create model
ln_model = lm(log(homelesscap) ~ log(filerate), data = data_nozero)
summary(ln_model)

plot_diagnostics(ln_model, x_var = data_nozero$filerate)
```

Although the average value of residuals seem to have moved much closer to 0, the heteroskedasticity is still bad. The coefficient for ln(filerate) indicates that on average a 1% increase in filerate is associated with a 4% decrease in homelesscap, significant at the 5% level.

The residual standard error for this model is lower (0.4656, previously 0.9464), also indicating improved fit. The $R^2$ is slightly lower, however, suggesting that filerate by itself is not a good predictor of homelesscap.

I will try adding some control variables to see if we can mitigate the omitted variable bias that is present. However, the control variables I have--red, blue and population density--may be plausibly correlated with filerate, so I am risking multicollinearity with the addition of these variables.

note: maybe talk about direction of bias as well

$~$

### Multivariable log-log
```{r}
mln_model = lm(log(homelesscap) ~ log(filerate) + red + blue + densityrank, data = data_nozero)
summary(mln_model)

plot_diagnostics(mln_model, x_var = data_nozero$filerate)
```
The variables in this model do a much better job of explaining the variability present in homelesscap, with an adjusted $R^2$ = 0.1267. While heteroskedasticity is still present, we can see from the QQ plot that the residuals seem relatively well-behaved otherwise. We can also see that the coefficient for filerate has decreased in both magnitude and statistical significance:
```{r}
confint(mln_model)
```
At 95% confidence, the coefficient of filerate could very well be 0; filerate may have no significant effect on homelesscap. The coefficient for red, however, is highly significant, and suggests that when a state is politically red, the rate of homelessness is 40% lower on average, holding population density and filing rate constant.

Knowing this, it is likely a good idea to take a closer look at state-fixed attributes.

note: check for multicollinearity

$~$

## State-fixed effects

```{r}
# scatter filerate on homelesscap, separating states with different politics and levels of population density
dlabs = c('1' = 'High Density', '0' = 'Low Density')
blabs = c('1' = 'Blue state', '0' = 'Red or Purple state')

base + geom_point(size = 0.8) +
  geom_smooth(method = 'lm', color = 'black', linewidth = .7) +
  geom_smooth(color = 'green', linewidth = .7) +
  facet_grid(rows = vars(dense), cols = vars(blue), 
             labeller = labeller(blue = blabs, dense = dlabs)) +
  labs(title = 'Homelessness rate v. eviction filings rate, by state characteristics') +
  ylab(hlabel) +
  xlab(flabel)
```

The relationship between filing rate and homelessness seems to vary wildly based on state characteristics. For this reason, a fixed effects model could work better for our data.

$~$

### Least squares dummy variable model

$$\hat{homelesscap}_{it} = \hat\beta_0(?) + \hat\beta_1{filerate}_{it} + \hat\gamma_{AK}{DAK}_i + ... + \hat\gamma_{WY}{DWY}_i + u_{it}$$

```{r}
lsdv_model = lm(homelesscap ~ filerate + factor(state_abbrev) - 1, data = data)
summary(lsdv_model)
```

note: not sure whether to omit intercept or not? bc i dont want to imply that
homeless = 0 when filerate = 0 but also would it cause multicollinearity? need to truncate output. i can keep a table of all the coefficients somewhere else

This model further reinforces the notion that filerate is not a good predictor for homelesscap; I am not sure whether $R^2$ can be used for this context but adj $R^2$ = 0.97--this model fits the data VERY well (to the point of being overfitted, possibly?), and most of the state dummy coefficients are highly significant, but the coefficient for filerate is once again insignificant. 
It looks like time-invariant state attributes may have a lot more to do with a state's rates of homelessness than a state's rate of eviction filings.

$~$

### Fixed-effects model (within)

```{r}
library(plm)
fe_model = plm(log(homelesscap) ~ log(filerate), data = data_nozero, index = c('state_abbrev', 'year'), model = 'within')
summary(fe_model)
```

notes: opposite results? filerate's coefficient now significant and positive, but adj r-sq is negative..?


## Discussion & conclusion

From these models (aside from last plm--need to further investigate), I conclude that rate of eviction filings does not significantly affect the rate of homelessness in a state.

Going forward, we could look more into our sample and data sources. we already know that the PIT counts are flawed measures, plus we could further investigate our panel dataset to see why it is unbalanced and whether we can fix that.

notes: 
-   correlation..??? covariance???
-   initial fe reg: explain overall, between and within variation
-   do a hausman test seems easy enough
    -   focus on within variation?? or time demean? maybe also a first differences test?
-   non linearity (how to combine non-linearity with fixed effects?) \*\*
-   EXPORT SOME TABLES?
-   find some tests to do to figure out best model ig. evaluation of fixed effects models
-   don't include all diagnostic graphs in this notebook. put them in a different script for those interested
